---
title: "Chicago Ridehail Open Data: Trips"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: true
    df_print: paged
---

```{r include=FALSE}
librarian::shelf(tidyverse, knitr, lubridate, ggplot2, 
                 reshape2, ggtext, ggthemes, httr, jsonlite, scales, RColorBrewer, RSocrata,
                 janitor, readxl, broom, scico)

knitr::opts_knit$set(echo = FALSE)
update_geom_defaults("line", aes(linewidth=1.5, alpha=1.0))
update_geom_defaults("point", aes(size=2, fill="Steel Blue", alpha=1.0))
# These values created 2022-06-03, key DriverProfile at https://data.cityofchicago.org/profile/edit/developer_settings
# api_key and api_key_secret apparently not used. Just the APP_TOKEN, but I'll keep them anyway
API_KEY_ID="blfvqesfen0k6tb0a2tm4uip1"
API_KEY_SECRET="4j1wla7fgnbt80obomq0a8po19ar9q9wwspdawakx17fa5jpr2"
TNC_TRIPS_ENDPOINT_OLD="https://data.cityofchicago.org/resource/m6dm-c72p.json"
TNC_TRIPS_ENDPOINT="https://data.cityofchicago.org/resource/n26f-ihde.json"
APP_TOKEN="UEuR3n1gEEHfMQf98QNFQG5yw"
```

```{r eval=FALSE, echo=FALSE}
# Test that access is working, and get the column names 
# The schema was updated in 2023: some judgement needed to merge the two sets
query=list(trip_id="0000000fb973b32717a335d3b7dd66deca2c5624")
res <- GET(TNC_TRIPS_ENDPOINT, add_headers(app_token=APP_TOKEN))
data = fromJSON(rawToChar(res$content))
names(data)
```
```{r eval=FALSE, echo=FALSE}
# w <- URLencode("trip_start_timestamp between '2019-10-01T0:00:00' and '2019-10-02T0:00:00'")
# w <- URLencode("trip_start_timestamp between '2022-12-01T00:00:00.000' and '2023-03-01T00:00:00.000'")
w <- URLencode("trip_start_timestamp between '2023-07-01T00:00:00.000' and '2023-08-01T00:00:00.000'")
url <- sprintf("%s?$where=%s&$order=trip_start_timestamp&$limit=5", TNC_TRIPS_ENDPOINT, w)
res <- GET(url, add_headers(app_token=APP_TOKEN))
tb <- as_tibble(fromJSON(rawToChar(res$content)))
tb
```

# Update the trip data with recent values

Using the first Thursday of each month, incrementally download data for select months and years. Save this and concatenate it with the trip_aggregates.csv file.

```{r average-cost-per-minute, eval=FALSE, echo=FALSE}
tb <- tibble(
  year=numeric(),
  month=numeric(),
  trip_count=numeric(), 
  fare_per_trip=numeric(),
  tip=numeric(),
  additional_charges=numeric(),
  trip_total_per_trip=numeric(),
  fare_per_minute=numeric(),
  trip_total_per_minute=numeric(),
  fare_per_mile=numeric(),
  miles_per_hour=numeric(), 
  minutes=numeric(), 
  miles=numeric(),
  .rows=NULL
)

for (year in c(2024)) {
for (month in c(07)){
 #for (year in c(2023)) {
#  for (month in c(10:11)){
    first_seven_dates <- seq(ymd(sprintf("%s-%s-01", year, month)),ymd(sprintf("%s-%s-07", year, month)),by="1 day")
    first_thursday <- first_seven_dates[wday(first_seven_dates, label=TRUE) == "Thu"]
    test_date <- format(first_thursday, "%Y-%m-%d")
    
    # These averages are rough: taking averages of different times and distances is not 
    # really correct. 
    q <- sprintf("SELECT 
            count(trip_id) as trip_count, 
            avg(fare) as fare_per_trip,
            avg(tip) as tip,
            avg(additional_charges) as additional_charges,
            avg(trip_total) as trip_total_per_trip,
            60.0 * avg(fare / trip_seconds) as fare_per_minute,
            avg(fare / trip_miles) as fare_per_mile,
            60.0 * avg(trip_total / trip_seconds) as trip_total_per_minute,
            avg(trip_total / trip_miles) as trip_total_per_mile,
            3600.0 * avg(trip_miles / trip_seconds) as miles_per_hour,
            avg(trip_seconds) / 60.0 as minutes,
            avg(trip_miles) as miles
          WHERE trip_seconds IS NOT NULL 
          AND trip_seconds > 0
          AND trip_miles > 0
          AND trip_start_timestamp between '%sT00:00:00.000' and '%sT23:59.000'", test_date, test_date)
    url <- URLencode(sprintf("%s?$query=%s", TNC_TRIPS_ENDPOINT, q))
    res <- GET(url, add_headers(app_token=APP_TOKEN))
    data <- fromJSON(rawToChar(res$content)) %>% as_tibble()
    tb <- tb %>% add_row(year=year,
                         month=month,
                         trip_count=as.numeric(data$trip_count),
                         fare_per_trip=as.numeric(data$fare_per_trip),
                         tip=as.numeric(data$tip),
                         additional_charges=as.numeric(data$additional_charges),
                         trip_total_per_trip=as.numeric(data$trip_total_per_trip),
                         fare_per_minute=as.numeric(data$fare_per_minute),
                         trip_total_per_minute=as.numeric(data$trip_total_per_minute),
                         fare_per_mile=as.numeric(data$fare_per_mile),
                         miles_per_hour=as.numeric(data$miles_per_hour),
                         minutes=as.numeric(data$minutes),
                         miles=as.numeric(data$miles)
                         )
    print(sprintf("%s-%s: %s", year, month, test_date))
  }
}  

tb <- tb %>% clean_names()

# Write it out as trip_aggregates_new just to avoid overwriting good data (trip_aggregates.csv) by mistake.
# If running this for a single month, append the results to trip_aggregates.csv
# If running the complete collection, which takes a long time, rename the file to 
# trip_aggregates.csv before continuing.
tb %>% write_csv('./data/trip_aggregates_new.csv', append=FALSE)
```


```{r echo=FALSE}
# Get the trip data into a tibble, called trips
trips <- read_csv("./data/trip_aggregates.csv", col_types="iiidddd") %>%
  mutate(date=as.Date(sprintf("%s-%s-01", year, month))) %>% 
  pivot_longer(c(trip_count, fare_per_trip, tip, additional_charges,
                 trip_total_per_trip,
                 fare_per_minute, trip_total_per_minute,
                 fare_per_mile, miles_per_hour, minutes, miles),
               names_to="category", values_to="measure"
               ) %>%
  filter(date > as.Date("2016-12-30"))
```

# Total number of trips

```{r echo=FALSE}
tb <- trips %>%
  filter(category == "trip_count")

p <- ggplot(data=tb,
            mapping=aes(x=date, y=measure, colour=category, fill=category))

p +
  geom_line() +
  # geom_point(shape=21, fill="white", colour="steelblue", size=3, stroke=2) +
  theme(legend.position="none") +
  scale_y_continuous(limits=c(0, NA), labels=label_number(scale_cut=cut_short_scale())) +
  labs(title="Ridehail trips in Chicago", 
       subtitle="One point per month, representing trips on the first Thursday of the month",
       x="Date",
       y="Daily trips") +
  scale_colour_brewer(palette="Dark2")
```

# Trip durations

```{r plot-chicago-trip-monthly-averages-drilldown, echo=FALSE}
tb <- trips %>%
  filter(category=="minutes")

p <- ggplot(data=tb, # %>% filter(category %in% c("FarePerMinute", "TripTotalPerMinute")),
            mapping=aes(x=date, y=measure, colour=category))

p +
  #facet_wrap(vars(category),
  #           scales = "free_y") +
  geom_line() +
 # geom_point(shape=21, fill="white", colour="steelblue", size=3, stroke=2) +
 # geom_line(colour="forestgreen", alpha=0.6) +
  theme(legend.position="none") +
  scale_y_continuous("", limits=c(0, NA), labels=label_number(scale_cut=cut_short_scale())) +
   ggtitle("Chicago ridehail aggregates: mean trip duration (minutes)", 
          subtitle="One point per month, representing the first Thursday of the month")+
  scale_colour_brewer(palette="Dark2")
```

# Trip durations and distances

With miles, minutes, amd miles per hour we should have a good overall view of fare duration and length.

```{r echo=FALSE}
tb <- trips %>%
    filter(category %in% c("minutes", "miles", "miles_per_hour"))
 
p <- ggplot(data=tb, # %>% filter(category %in% c("FarePerMinute", "TripTotalPerMinute")),
            mapping=aes(x=date, y=measure, colour=category, fill=category))

p +
  facet_wrap(vars(category),
             scales = "free_y") +
  # geom_point() +
  geom_line() +
  theme(legend.position="none") +
  scale_y_continuous("", limits=c(0, NA), labels=label_number(scale_cut=cut_short_scale())) +
  ggtitle("Chicago ridehail aggregates", 
          subtitle="One point per month, representing the first Thursday of the month") +
  scale_colour_brewer(palette="Dark2")
```

# Trip fares

Fare per mile, fare per minute, and minutes per fare. With these three we should have a basic overall-fare-per-minute metric (or, equivalently, overall-fare-per-mile).

Next step - calculate that.


```{r plot-chicago-trip-monthly-averages, echo=FALSE}
tb <- trips %>%
  filter(category %in% c("fare_per_minute", "fare_per_mile", "trip_total_per_minute"))

p <- ggplot(data=tb,
            mapping=aes(x=date, y=measure, colour=category, fill=category))

p +
  facet_wrap(vars(category),
             scales = "free_y") +
  # geom_point() +
  geom_line() +
  theme(legend.position="none") +
  scale_y_continuous("", limits=c(0, NA), labels=label_number(scale_cut=cut_short_scale())) +
  labs(title="Chicago ridehail aggregates", 
          subtitle="One point per month, representing the first Thursday of the month") +
  scale_colour_brewer(palette="Dark2")
```

# Fare total per trip

Multiply the "trip_total_per_minute" by "minutes" to get a "trip total" fare.

```{r fare-total-per-trip, echo=FALSE}
tb <- trips %>% 
  filter(category %in% c("fare_per_trip", "tip", "additional_charges", "trip_total_per_trip")) %>%
  pivot_wider(names_from = category, values_from = measure) %>%
  mutate(fare_without_tip = trip_total_per_trip - tip) %>%
  pivot_longer(cols = c(fare_per_trip, tip, additional_charges, trip_total_per_trip, fare_without_tip), 
               names_to = "category", values_to = "measure") %>%
  filter(category %in% c("fare_without_tip"))

p <- ggplot(data=tb, 
            mapping=aes(x=date, y=measure, colour=category))

p +
  #facet_wrap(vars(category),
  #           scales = "free_y") +
  geom_line() +
  #geom_point(shape=21, fill="white",  size=3, stroke=2) +
  scale_y_continuous(limits=c(0, NA), labels=label_number(scale_cut=cut_short_scale())) +
  labs(title="Chicago ridehail aggregates: average fare per trip",
       subtitle="One point per month, representing the first Thursday of the month",
       x="Date",
       y="Fare per trip ($)") +
  theme(legend.position=c(0.2, 0.85), legend.title=element_blank()) +
    scale_colour_brewer(palette="Dark2",
                       labels=c("Fare per trip (excluding tip)"))
```

# Fares and gasoline prices

The US Energy Information Administration publishes gasoline prices in cities, including Chicago. Here is a chart, downloaded from its open data site [here](https://www.eia.gov/dnav/pet/hist/LeafHandler.ashx?n=pet&s=emm_epmrr_pte_yord_dpg&f=w).

(To download, go to the site and click "Download Data (xls file)")

```{r chicago-gas, echo=FALSE}
# Read the gas data from the spreadsheet, and then calculate average monthly prices
# to match the Chicago ridehail data
gas <- read_xls("./data/EMM_EPMRR_PTE_YORD_DPGw.xls", 
                col_names=c("date", "price"),
                col_types = c("date", "numeric"),
                sheet="Data 1", skip=3) %>% 
  group_by(month=lubridate::floor_date(date, "month")) %>% 
  summarise(price=mean(price)) %>%
  filter(month >= as.Date("2019-01-01")) %>%
  ungroup() %>%
  mutate(month=as.Date(month)) %>%
  mutate(category="gas_price") %>%
  rename(measure=price)

tb <- trips %>%
  select(c(date, category, measure)) %>%
  rename(month=date) %>%
  union(gas) %>%
  filter(category %in% c("trip_total_per_minute", "gas_price")) %>%
  filter(month > as.Date("2016-12-30"))

p <- ggplot(tb, map=aes(x=month, y=measure, colour=category))

p +
  geom_line() +
 # geom_point(shape=21, fill="white", size=3, stroke=2) +
  labs(x="Date",
       y="Price") +
  scale_x_date(date_breaks="6 month", date_labels="%b %Y") +
  scale_y_continuous(limits=c(0, NA)) +
  theme (legend.position=c(0.2, 0.8), 
        legend.title=element_blank()) +
  scale_colour_brewer(palette="Dark2",
                       labels=c("Gas price ($/gal)", "Trip total fare ($/min)"))
```

# Hourly trends during the day (first Thursday in February)

```{r feb6-collection, eval=FALSE, echo=FALSE}
tb <- tibble(
  year=numeric(),
  month=numeric(),
  hour=numeric(),
  trip_count=numeric(), 
  fare_per_minute=numeric(),
  trip_total_per_minute=numeric(),
  fare_per_mile=numeric(),
  miles_per_hour=numeric(), 
  minutes=numeric(), 
  miles=numeric(),
  .rows=NULL
)

# First Thursday in the month
year <- 2024
month <- 02
day <- 01
test_date <- sprintf("%04i-%02i-%02i", year, month, day) # Match Toronto Feb 6, 2020 

for (hour in c(0:23)) {
  # These averages are rough: taking averages of different times and distances is not 
  # really correct. 
  q <- sprintf("SELECT 
          count(trip_id) as trip_count, 
          60.0 * avg(fare / trip_seconds) as fare_per_minute,
          60.0 * avg(trip_total / trip_seconds) as trip_total_per_minute,
          avg(fare / trip_miles) as fare_per_mile,
          3600.0 * avg(trip_miles / trip_seconds) as miles_per_hour,
          avg(trip_seconds) / 60.0 as minutes,
          avg(trip_miles) as miles
        WHERE trip_seconds IS NOT NULL 
        AND trip_seconds > 0
        AND trip_miles > 0
        AND trip_start_timestamp between '%sT%02i:00:00.000' and '%sT%02i:59.999'", test_date, hour, test_date, hour)
  url <- URLencode(sprintf("%s?$query=%s", TNC_TRIPS_ENDPOINT, q))
  res <- GET(url, add_headers(app_token=APP_TOKEN))
  data <- fromJSON(rawToChar(res$content)) %>% as_tibble()
  tb <- tb %>% add_row(year=year,
                       month=month,
                       hour=hour,
                       trip_count=as.numeric(data$trip_count),
                       fare_per_minute=as.numeric(data$fare_per_minute),
                       trip_total_per_minute=as.numeric(data$trip_total_per_minute),
                       fare_per_mile=as.numeric(data$fare_per_mile),
                       miles_per_hour=as.numeric(data$miles_per_hour),
                       minutes=as.numeric(data$minutes),
                       miles=as.numeric(data$miles)
                       )
  # print(sprintf("%s:%02i", test_date, hour))
}  

# Append this csv to trip_aggregates_hourly.csv for the first Thursday in February of each year
# (see next cell)
# tb %>% write_csv(sprintf('./data/trip_aggregates_%s.csv', test_date), append=FALSE)
```

```{r plot-chicago-hourly-averages, echo=FALSE}
year <- 2024
month <- 2
day <- 1
# test_date <- sprintf("%04i-%02i-%02i", year, month, day) # Match Toronto Feb 6, 2020
test_date <- sprintf("%04i-%02i-%02i", year, month, day) # Match Toronto Feb 6, 2020 

# trip_aggregates_hourly.csv includes values from trip_aggregates_%s.csv,
# mainly for Feb 6.
tb <- read_csv(sprintf("./data/trip_aggregates_hourly.csv"), col_types="cciiddddd") %>%
  pivot_longer(c(trip_count, fare_per_minute, trip_total_per_minute, fare_per_mile, miles_per_hour, minutes, miles),
               names_to="category", values_to="measure"
               ) %>%
  filter(category=="trip_count")

p <- ggplot(data=tb, mapping=aes(x=hour, y=measure, colour=year, fill=year, label=category))

p +
  # facet_wrap(vars(category), scales = "free_y") +
  geom_line() +
  # geom_point(shape=21, fill="white", size=3, stroke=2) +
  # theme(legend.position="none") +
  scale_y_continuous(labels=label_number(scale_cut=cut_short_scale())) +
  labs(title=sprintf("Chicago ridehail: trip distribution on the first Thursday in February"),
       subtitle="One point per hour",
       x="Hour of the day",
       y="Number of trips") +
  scale_colour_brewer(palette="Dark2")
```

