---
title: "Chicago Ridehail Open Data: Drivers"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: true
  number_sections: true
  df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE, echo=FALSE}
# install.packages("librarian")
librarian::shelf(tidyverse, ggplot2, knitr, reshape2, ggtext, jsonlite, stringr, lubridate, scales, ggthemes, RColorBrewer, viridis, vroom, janitor)

knitr::opts_knit$set(echo = FALSE)
```

```{r, read-data, echo=FALSE}
# The N_ columns are integer count of months from 2011-08.
# csv_file="driver_sample.csv"
# Reading and preparing the drivers CSV takes 20 -- 30 seconds
csv_file="./data/Transportation_Network_Providers_-_Drivers.csv"
drivers <-vroom(csv_file, col_types=cols(
  MONTH_REPORTED=col_date(format="%Y-%m"),
  DRIVER_START_MONTH=col_date(format="%Y-%m"),
  CITY=col_character(),
  STATE=col_character(),
  ZIP=col_character(),
  NUMBER_OF_TRIPS=col_integer(),
  MULTIPLE_TNPS=col_logical()
  )) %>%
  mutate(
    N_MONTH_REPORTED=as.integer(round((MONTH_REPORTED - as.Date("2011-08-01"))/(365.25/12))),
    N_DRIVER_START_MONTH=as.integer(round((DRIVER_START_MONTH - as.Date("2011-08-01"))/(365.25/12))),
    DRIVER_LONGEVITY=N_MONTH_REPORTED - N_DRIVER_START_MONTH,
    FACTOR_MONTH_REPORTED=as.factor(MONTH_REPORTED)
  ) %>%
  as_tibble() %>%
  clean_names()
```

# Data source

Chicago publishes three data sets about Transportation Network Providers operations in the city. This notebook uses the [Drivers](https://data.cityofchicago.org/Transportation/Transportation-Network-Providers-Drivers/j6wf-834c) data set. All three are updated monthly. I download the Drivers set as a CSV file; the Trips data set is too big for that, and (in other notebooks) I query it from the notebook.

After download, remove  some obvious gaps and errors (like drivers with a start date before 2011, or drivers with a start date after the month reported). 


In this notebook, "drivers" = "active drivers", which is to say drivers with a number of trips > 0 in a given month. 


```{r, clean-data, echo=FALSE}
# print(sprintf("Before cleaning: %s", count(drivers)))
active_drivers <- drivers %>% 
  filter(!is.na(number_of_trips), 
         !is.na(driver_start_month),
         driver_start_month >= as.Date("2011-08-01"),
         driver_start_month <= month_reported,
         number_of_trips > 0
        )
# print(sprintf("Active drivers: %s", count(active_drivers)))
```

# Trips per month

Start with an overview of trip volume over time. 

```{r, trips-per-month, echo=FALSE}
tb <- active_drivers %>% 
  select(c(month_reported, number_of_trips)) %>%
  group_by(month_reported) %>%
  replace_na(list(number_of_trips=0)) %>% 
  summarize(trips=sum(number_of_trips), drivers=n()) %>%
  ungroup() %>%
  mutate(trips_per_driver = trips/drivers) %>%
  pivot_longer(cols=c(trips, drivers, trips_per_driver), names_to="measure", values_to="value")
    

p <- ggplot(data= tb %>% 
              filter(measure=="trips") %>% 
              filter(month_reported > as.Date("2016-12-30")), 
            aes(x=month_reported, y=value, colour=measure))
p + 
  #facet_wrap(vars(measure), scales="free") +
  geom_point(size=2) +
  labs(title="TNC trips per month in Chicago",
         x="Year",
         y="Trips per month") + 
  scale_y_continuous(limits=c(0, NA), 
                     labels=label_number(scale_cut=cut_short_scale())) +
  theme(legend.position="none") +
  scale_colour_tableau(palette="Tableau 10")
```

**Discussion**

There are three periods:

1. Expansion until about 9M trips per month at the beginning of 2018
2. Flatlining at 9M trips per day during 2018 and 2019
3. Covid-driven collapse in early 2020 followed by steady recovery during
2021, so in mid-2022 the number of trips is back to about 2/3 of pre-Covid levels.

For comparison with other cities, 9M trips per month is equivalent to:
- 300,000 per day, or
- 12000 per hour, or
- 200 per minute.


# Driver population

The trend for active drivers parallels that for trips.

The trends for all drivers (including those with no trips) have some similarities and differences to the trips:

1. Expansion until early 2017. 
2. During the first half of 2017 the number of drivers falls off and levels out at about 70K. This is while the number
of trips is still increasing.
3. Flatlining during 2017-H2, 2018, 2019.
3. A slower and less complete Covid-driven drop-off of drivers during 2020 than trips, followed by some recovery during 2021.

Trips and drivers are now back to about 0.6 of the pre-Covid numbers.

Regarding the outliers in Q2 2018, there is a note on the City of
Chicago web site:

> [S]ome vehicle records for Q2 2018 (April-June) were reported for the
> quarter as a whole, rather than for individual months. For purposes of
> this dataset, those records have been assigned to 2018-06 (June).
> Therefore, some caution in interpreting this month and the quarter as
> a whole is advised.

```{r echo=FALSE}
tb <- active_drivers %>% 
  select(c(month_reported)) %>%
  group_by(month_reported) %>%
  summarize(drivers=n()) %>%
  ungroup() %>%
  mutate(dummy = 'x')

p <- ggplot(data=tb %>% filter(month_reported > as.Date("2016-12-30")),
            aes(x=month_reported, y=drivers, colour=dummy))
p + 
  geom_point(size=2) +
  labs(title="TNC active drivers in Chicago",
       subtitle="Active drivers are those who drive at least one trip in the month",
       x="Date",
       y="Active drivers") +
  scale_x_date(date_breaks = "1 year", date_labels = "%b %Y") +
  scale_y_continuous(limits=c(0, NA), labels=label_number(scale_cut=cut_short_scale())) +
  theme(legend.position="none") +
  scale_color_tableau(palette="Tableau 10")
```

**Discussion**

The basic pattern is the same as that for trips. The Covid plummet is a bit less complete, although still dramatic.


## Drivers, by hours driven

There is no one Uber driver. To understand how Uber operates, it helps to know what the range of driver participation is. Some drive only occasionally, some drive full time.

```{r, driver-distribution, echo=FALSE}
selected_month <- as.Date("2023-02-01")
current_month <- 
tb <- active_drivers %>%
  filter(month_reported == selected_month) %>%
  mutate(dummy = 'x')

p <- ggplot(data=tb, 
            mapping=aes(x=number_of_trips, fill=dummy)) 

p + 
  geom_histogram(alpha=0.6, binwidth = 50, position=position_nudge(25), colour="white") +
  labs(title=sprintf("Distribution of drivers, by number of trips in the month, for %s", selected_month),
        subtitle="The large number of inactive drivers with no trips at all are not shown.",
       x="Trips",
       y="Drivers") +
  theme(legend.position="none") +
  scale_fill_tableau(palette="Tableau 10")
```

## Categorizing drivers

The Chicago data set does not provide driver times on the platform. The nearest we get is the number
of trips in the month.

Using an average of 20 minutes per trip (from the Trips data set) and a utilization rate of 66% (a generous estimate from Fehr & Peers) gives two trips per hour. A 30 hour week translates into 60 trips, which is about 250 trips per month. 

Any attempt to divide the distribution into categories is a bit arbitrary, but this section shows a few categorization schemes.


```{r, driver-categories, echo=FALSE}
driver_category_seattle <- function(trips){
  if (trips <= 40) {
    return("Casual")
  } else if (between(trips, 41, 160)) {
    return("Committed casual") 
  } else if(between(trips, 161, 240)) {
    return("Part-time")
  } else if (trips > 240) {
    return("Full-time")
  } else {
    return("NA")
  }
}

driver_category_seattle_v <- Vectorize(driver_category_seattle)

driver_category_2 <- function(trips){
  if (trips <= 40) {
    return("1")
  } else if (between(trips, 41, 160)) {
    return("2") 
  } else if(between(trips, 161, 240)) {
    return("3")
  } else if (between(trips, 241, 320)) {
    return("4")
  } else {
    return("5")
  }
}
driver_category_2_v <- Vectorize(driver_category_2)

full_time_trips = 250
driver_category_coarse <- function(trips){
  if (trips <= full_time_trips) {
    return ("Part time")
  } else {
    return ("Full time")
  }
}
driver_category_coarse_v <- Vectorize(driver_category_coarse)

calendar_month <- "06"
test_months = c(as.Date(sprintf("2015-%s-01", calendar_month)),
                as.Date(sprintf("2016-%s-01", calendar_month)),
                as.Date(sprintf("2017-%s-01", calendar_month)),
                as.Date(sprintf("2018-%s-01", calendar_month)),
                as.Date(sprintf("2019-%s-01", calendar_month)),
                as.Date(sprintf("2020-%s-01", calendar_month)),
                as.Date(sprintf("2021-%s-01", calendar_month)),
                as.Date(sprintf("2022-%s-01", calendar_month))
                )

measure_label <- c("driver_percent"= "Drivers", "trip_percent" = "Trips")
```

### Seattle classification

First, The Seattle classification of Hyman et al to map that onto the Chicago data. The
categories are:

| Category | Trips      | Estimated hours / week |
|----------|------------|------------------------|
| 1        | \< 40      | \< 5                   |
| 2        | 41 -- 160  | 5 -- 20                |
| 3        | 161 -- 240 | 20 -- 30               |
| 4        | 241 -- 320 | 30 -- 40               |
| 5        | \> 320     | \> 40                  |

The proportion of Casual drivers fell off quickly from 2016 to 2018, and
the number of part-time (160-240) and full-time drivers rose. Since
then, the composition has changed only a little.


```{r category-coarse, echo=FALSE}
category <- driver_category_coarse_v

tb <- active_drivers %>% 
  select(c(month_reported, number_of_trips)) %>%
  # filter(month_reported %in% test_months) %>%
  mutate(driver_category = category(number_of_trips)) %>%
  group_by(month_reported, driver_category) %>%
  summarise(drivers=n(), trips=sum(number_of_trips)) %>%
  mutate(driver_percent=proportions(drivers) * 100,
         trip_percent=proportions(trips) * 100) %>%
  ungroup %>%
  pivot_longer(cols=c(driver_percent, trip_percent),
               names_to="measure",
               values_to="value") %>%
  arrange(month_reported, driver_category)

# p <- ggplot(data=driver_categories, 
#            mapping=aes(x=driver_category, y=driver_percent, fill=factor(month_reported)))

 p <- ggplot(data=tb, 
            mapping=aes(x=month_reported, y=value, 
                        fill=driver_category,
                        label=measure))
 
p + 
  facet_wrap(~measure, labeller=labeller(measure=measure_label)) +
  # geom_line(size=1) +
  geom_area(alpha=0.7) + 
  # geom_point(size=3) +
  labs(title="Chicago ridehail driver hours-of-work categories",
       subtitle=sprintf("Full timers (over %s trips per month) make up a small proportion of all drivers,\nbut deliver a larger proportion of trips", full_time_trips),
       x="Month",
       y="Percent",
       colour="Driver category") +
  scale_y_continuous(breaks=waiver(), n.breaks=6) +
  theme(legend.position="top") +
  # theme(panel.grid.major=element_line(colour="grey")) +
  guides(fill=guide_legend(title="Driver category")) +
  scale_fill_tableau(palette="Tableau 10") 
```

Here are the values for the most recent month:

```{r full-time-this-month}
tb %>% filter(month_reported==max(month_reported))
```

**Discussion**

Despite the changes over the last five years, including the maturing of the Uber platform and Covid, the 
driver population has remained surprisingly consistent. Full time drivers (over 275 trips per month) make up only about one in six drivers, but deliver over 40% of all trips.

Full-timers and part-timers are both essential to Uber and Lyft's ability to deliver the volume of trips they handle. But these two groups do not always share the same interests.

## Mean and median number of trips per driver

Although there is no "one type of Uber driver", averages may still be useful to track trends.

The average number of trips per driver increased during 2017 (as measured both by
median and mean): the number of drivers stayed roughly the same
while the number of trips continued to grow. In 2018 and 2019 there was
a roughly steady state, with an average of about 125 trips per month
(about 16 hours per week), a number that has been restored during 2022.


```{r trips-per-driver, echo=FALSE}
tb <- active_drivers %>%
  select(c(month_reported, number_of_trips)) %>%
  group_by(month_reported) %>%
  summarise(mean=mean(number_of_trips),
            median=median(number_of_trips)) %>%
  pivot_longer(cols=c(mean, median), names_to="average", values_to="trips")

p <- ggplot(data=tb, aes(x=month_reported, y=trips, colour=average))
p + 
  geom_point(size=2) +
  labs(title="Average trips per active driver", 
        subtitle="Showing both mean and median number of trips",
        x="Date",
        y="Trips/month") +
  scale_x_date(date_labels="%b %Y") +
  scale_y_continuous(limits=c(0, NA), labels=label_number(scale_cut=cut_short_scale())) +
  theme(legend.position="top") +
  scale_color_tableau(palette="Tableau 10")
```

# Driver longevity and churn

## Driver population by longevity

The first chart takes a specific month and shows the distribution of
drivers, by longevity, on the platform in that month.

```{r driver-longevity, echo=FALSE}
reference_month = as.Date("2023-02-01")

tb <- active_drivers %>% 
  select(c(month_reported, driver_longevity)) %>%
  filter(month_reported==reference_month) %>% 
  filter(driver_longevity > 0) %>%
  group_by(driver_longevity) %>% 
  summarize(Drivers=n()) %>%
  ungroup() %>%
  mutate(dummy='x')

mean_start <- active_drivers %>% 
  filter(month_reported==reference_month) %>% 
  summarize(start=mean(driver_start_month))

median_start <- active_drivers %>% 
  filter(month_reported==reference_month) %>% 
  summarize(start=median(driver_start_month))

p <- ggplot(data=tb, aes(x=driver_longevity, y=Drivers, colour=dummy))

p + geom_point(size=2) +
  labs(title=sprintf("Longevity of active drivers reported in %s", reference_month),
       subtitle=sprintf("Mean start time=%s; mean longevity=%s months; median longevity=%s months", 
                        mean_start$start[1], 
                        round((reference_month - mean_start$start[1])/30),
                        round((reference_month - median_start$start[1])/30)
                        ),
       x="Driver longevity (months)") +
  theme(legend.position="none") +
  scale_color_tableau(palette="Tableau 10")
```

It can be easier to interpret this if we lump drivers into new or long-term workers, based on 
whether they have been on the platform for more or less than a year.

```{r driver-longevity-category, echo=FALSE}
driver_longevity_category <- function(driver_longevity){
  if (driver_longevity <= 12) {
    return ("A year or less")
  } else {
    return ("More than a year")
  }
}
driver_longevity_category_v <- Vectorize(driver_longevity_category)

category <- driver_longevity_category_v

measure_label <- c(
  "driver_percent"= "Drivers",
  "trip_percent" = "Trips"
)

tb <- active_drivers %>% 
  select(c(month_reported, driver_longevity, number_of_trips)) %>%
  #filter(month_reported %in% test_months,
  #       driver_longevity > 0) %>%
  mutate(driver_category = category(driver_longevity)) %>%
  group_by(month_reported, driver_category) %>%
  summarise(drivers=n(), trips=sum(number_of_trips)) %>%
  mutate(driver_percent=proportions(drivers) * 100,
         trip_percent=proportions(trips) * 100) %>%
  ungroup %>%
  pivot_longer(cols=c(driver_percent, trip_percent),
               names_to="measure",
               values_to="value")

p <- ggplot(data=tb, aes(x=month_reported, y=value, colour=driver_category, label=measure))

p + 
  facet_wrap(~measure, labeller=labeller(measure=measure_label)) +
  #geom_point(size=2) +
  geom_line(size=1) +
  labs(title="Chicago ride hail driver longevity categories",
       subtitle="Drivers with a year or less on the platform still make up a large proportion of drivers\n and provide a large proportion of trips", 
       caption="Ride hailing was still new and fast-growing in 2016, so there were few long-term drivers.\nSince then, the churn of incoming and leaving drivers has evened out.",
       x="Time",
       y="Percent",
       colour="Driver longevity") +
  scale_y_continuous(breaks=waiver(), n.breaks=6) +
  theme(legend.position="top") +
  guides(fill=guide_legend(title="Driver category")) +
  scale_color_tableau(palette="Tableau 10")
```

Both mean and median longevity (below) confirm the observation that the driver population
has been on the platform for longer, over time, albeit with a major
Covid interruption.

The Chicago data does not allow us to see whether drivers have gaps in
their work: we just have a first month they are recognized and the
reporting month.

The typical driver has now been on the platform for almost a year.
Drivers who drive more trips have, on average, been on the platform just
a little longer.

```{r mean-longevity, echo=FALSE}
tb <- active_drivers %>% 
  select(c(month_reported, driver_longevity)) %>%
  group_by(month_reported) %>% 
  summarize(mean_longevity=mean(driver_longevity)) %>%
  ungroup() %>%
  mutate(dummy='x')
  
p <- ggplot(data=tb,
            aes(x=month_reported, y=mean_longevity, 
                colour=dummy))

p + geom_point(size=2) +
  #geom_smooth() +
  labs(title="Chicago active driver mean longevity",
       x = "Date",
       y = "Mean active driver longevity (months)") +
  scale_x_date(date_label="%b %Y") +
  theme(legend.position="none") +
  scale_colour_tableau(palette="Tableau 10")
```

```{r median-longevity, echo=FALSE}
tb <- active_drivers %>% 
  filter(number_of_trips > 240) %>%
  select(c(month_reported, driver_longevity)) %>%
  group_by(month_reported) %>% 
  summarize(median_longevity=median(driver_longevity)) %>%
  ungroup() %>%
  mutate(dummy='x')
  
p <- ggplot(data=tb,
            aes(x=month_reported, y=median_longevity, colour=dummy))

p + geom_point(size=2) +
  labs(title="Median active driver longevity over time",
       x="Time",
       y="Median active driver longevity (months)") +
  theme(legend.position="none") +
  scale_colour_tableau()
```


## Driver churn

The next chart takes a given starting month, and looks at how many
drivers are left over time.

```{r churn-start, echo=FALSE}
reference_month = as.Date("2018-01-01")

tb <- active_drivers %>% 
  filter(driver_start_month == reference_month) %>% 
  filter(month_reported > reference_month) %>% 
  group_by(month_reported) %>% 
  summarize(drivers=n()) %>%
  ungroup() %>%
  mutate(dummy='x')

mean_expiry_month <- drivers %>% 
  filter(driver_start_month==reference_month) %>% 
  summarize(expiry=mean(month_reported)) %>% 
  pull(expiry)

p <- ggplot(data=tb, aes(x=month_reported, y=drivers, colour=dummy))

p + 
  geom_point(size=2) +
  geom_smooth(method="gam", se=TRUE) +
  labs(title=sprintf("Active drivers starting in %s, still on the platform at a later time", reference_month),
       subtitle=sprintf("Mean time on platform = %s days", round(mean_expiry_month-reference_month)),
       x="Month",
       y="Drivers") +
  theme(legend.position = "none") +
  scale_colour_tableau(palette="Tableau 10")
```


Now see if we can say "how many quit over their first year"

```{r driver-churn-facet, echo=FALSE}
# Cannot compute this for dates less than a year old
tb <- active_drivers %>% 
  filter(
    driver_start_month > as.Date("2016-01-01"),
    driver_start_month <= as.Date("2022-02-01"),
    month_reported == driver_start_month + months(1) |
    month_reported == driver_start_month + years(1)) %>%
  group_by(driver_start_month, month_reported) %>%
  summarise(drivers=n(),
            trips=sum(number_of_trips)) %>%
  ungroup() %>%
  arrange(driver_start_month, month_reported) %>%
  mutate(driver_survival = drivers / lag(drivers),
         trip_survival = trips / lag(trips)) %>%
  filter(driver_survival < 1 & trip_survival < 1) %>%
  pivot_longer(cols=c("driver_survival", "trip_survival"), names_to="measure", values_to="value")

measure_label <- c(
  "driver_survival"= "Drivers",
  "trip_survival" = "Trips"
)

p <- ggplot(data=tb, aes(x=driver_start_month, y=value, colour=measure, label=measure))

p + 
  facet_wrap(~measure, scale="free", labeller=labeller(measure=measure_label)) +
  ylim(0, NA) +
  geom_point(size=2) +
  # geom_smooth(method="gam", se=TRUE) +
  labs(title="Chicago ridehail driver churn",
    subtitle="Fraction of drivers still active after a year, and fraction of trips provided by those drivers",
    caption="Over two-thirds of drivers quit within a year. Over half of all trips are provided by drivers who last less than a year.",
    x="Date",
    y="Fraction",
    colour="Quantity") +
  theme(legend.position = "none") +
  scale_colour_tableau(palette="Tableau 10")
```

## Trip distribution

In a given month, how many trips are given by drivers with a residency
of X months?

```{r trip-longevity, echo=FALSE}
reference_months = list(as.Date("2016-02-01"), 
                        as.Date("2017-02-01"), 
                        as.Date("2018-02-01"), 
                        as.Date("2019-02-01"),
                        as.Date("2020-02-01"), 
                        as.Date("2021-02-01"),
                        as.Date("2022-02-01"),
                        as.Date("2023-02-01")
                        )

tb <- active_drivers %>% 
  filter(month_reported %in% reference_months) %>% 
  filter(driver_start_month < month_reported) %>% 
  select(c(factor_month_reported, driver_longevity, number_of_trips)) %>%
  group_by(factor_month_reported, driver_longevity) %>%
  summarise(trips=sum(number_of_trips)) %>%
  replace_na(list(trips=0)) %>%
  mutate(trip_fraction=prop.table(trips))

p <- ggplot(data=tb,
            aes(x=driver_longevity, y=trip_fraction, colour=factor_month_reported))
p + geom_point(size=1, alpha=0.8) +
  ggtitle("The evolution of active driver longevity") +
  scale_x_continuous("Active driver longevity (months)", limits=c(0,24), labels=label_number(scale_cut=cut_short_scale())) +
  scale_y_continuous("Trip Fraction", labels=label_number(scale_cut=cut_short_scale())) +
  geom_smooth(se=FALSE) +
  theme(legend.position = c(0.8, 0.7)) +
  scale_colour_tableau()
```

Now do this over all months, not just an individual month, summing over
the longevity of the driver.

```{r longevity, echo=FALSE}
tb <- active_drivers %>% 
  filter(!is.na(driver_start_month), driver_longevity>0) %>% 
  select(c(driver_longevity, number_of_trips)) %>% 
  group_by(driver_longevity) %>% 
  summarise(trips=sum(number_of_trips))

mean_longevity <- tb %>% 
  summarise(mean=sum(driver_longevity * trips)/sum(trips)) %>% 
  pull(var=mean)

p <- ggplot(data=tb,
            aes(x=driver_longevity, y=trips))

p + 
  geom_point(size=2, colour="steelblue") +
  geom_line(linetype="dashed", colour="steelblue") +
  ggtitle("Number of trips, by active driver longevity",
          subtitle=sprintf("Mean longevity = %s months", round(mean_longevity))) +
  xlab("Active driver longevity on platform (months)") +
  ylab("Trips") +
  scale_y_continuous(labels=label_number(scale_cut=cut_short_scale())) +
  geom_vline(xintercept = mean_longevity, linetype="dashed", 
                color = "steelblue", alpha=0.6, size=0.5) +
  scale_colour_tableau()
```

## New starts and quits, by month

First let's do starts, as that is easy.

The data point for Jan 2016 was a BIG outlier, and is excluded.

```{r echo=FALSE}
tb_starts <- active_drivers %>% 
  filter(driver_start_month==month_reported) %>%
  filter(driver_start_month != as.Date("2016-01-01")) %>%
  select(c(driver_start_month)) %>% 
  group_by(driver_start_month) %>% 
  summarise(starts=n())%>%
  arrange(driver_start_month)

p <- ggplot(data=tb_starts,
            aes(x=driver_start_month, y=starts))
p + geom_point(size=2, colour="steelblue") +
  ggtitle("Number of drivers starting each month") +
  xlab("Month") +
  scale_y_continuous("Drivers starting", labels=label_number(scale_cut=cut_short_scale())) +
  scale_colour_tableau()
```

```{r echo=FALSE}
tb <- drivers %>% 
  select(c(month_reported)) %>%
  group_by(month_reported) %>%
  summarise(drivers=n()) %>%
  arrange(month_reported) %>%
  mutate(diff = drivers - lag(drivers))

tb_long <- tb %>% 
  left_join(tb_starts, by=c("month_reported"="driver_start_month")) %>%
  mutate(quits=starts-diff) %>% 
  select(c(month_reported, drivers, starts, quits)) %>%
  pivot_longer(cols=c(starts, quits),
                    names_to=c("category"))

p <- ggplot(data=tb_long, aes(x=month_reported, y=value, colour=category))
p + 
  geom_point(size=2) +
  xlab("Month") +
  scale_y_continuous("Drivers starting or quitting", limits=c(-1000, 10000), labels=label_number(scale_cut=cut_short_scale())) +
  # scale_colour_brewer(palette = "Dark2") +
  scale_colour_tableau()
```
