---
title: "Uber Upfront Pay: Evidence from Driver Screenshots"
# subtitle: "(preliminary)"
date: last-modified
author: "Rideshare Drivers Association of Ontario"
format: 
  revealjs:
    # logo: whatever.png
    footer: "Copyright © 2024 Rideshare Drivers Association of Ontario"
---

Following Uber's introduction of "Upfront pricing for drivers" in Ontario on October 8, 2024, rideshare drivers have been sending screenshots of trip offers and trip earnings to The Rideshare Drivers Association of Ontario. This set of charts captures some of what those screenshots show. 

The charts compare "upfront" offers to what the previous "rate card" pay would have been for a trip of that time and distance. The rate card pay is calculated from the per km, per minute, and base fare values for the service (UberX, UberXL etc), with 25\% removed for Uber's take.

The first two charts show the same overall averages in two different ways, as a scatterplot and as a histogram.

---

```{r}
#| label: setup
#| include: FALSE

librarian::shelf(tidyverse, googlesheets4, janitor)
ss = "1_jOPbHF2T3tm7a3jM3T0giYQgeqi7isB9JIH4hmtnqE"
ss_id = as_sheets_id(ss)
url = sprintf("https://docs.google.com/spreadsheets/d/%s/edit?gid=1094854580#gid=1094854580", ss)
tb_ref <- read_sheet(ss=ss_id, range="Reference Values!A2:I5") %>% 
  column_to_rownames(var="Metric") %>% 
  t() %>%
  as_tibble(rownames=NA) %>%
  rownames_to_column(var="trip_type") %>%
  clean_names()
tb <- read_sheet(ss=ss_id, 
                 sheet="Form Responses 1",
                 na = c("0", "0.0", "0.00")) %>% 
  clean_names() %>%
  rename("upfront_offer" = "upfront_offer_amount_enter_0_if_not_shown",
         "estimated_time" = "estimated_time_total_of_both_pickup_and_drop_off_mins",
         "estimated_distance" = "estimated_distance_total_of_both_pickup_and_drop_off_distance_km",
         "actual_earnings" = "actual_fare_payment_earnings_not_including_hst_enter_0_if_not_shown",
         "actual_time" = "actual_trip_time_pickup_and_drop_off_time_minutes_seconds_enter_0_if_not_shown",
         "actual_distance" = "actual_trip_distance_pickup_and_drop_off_distance_km_enter_0_if_not_shown",
         "image_file" = "image_file_name_no_extension_needed",
         "tip" = "tips_enter_0_if_not_shown"
         ) %>% 
  distinct(upfront_offer, estimated_time, estimated_distance, actual_earnings, actual_time, actual_distance, .keep_all = TRUE) %>%
  mutate(estimated_time_decimal = round(estimated_time) + (estimated_time %% 1) * 100 / 60,
         actual_time_decimal = round(actual_time) + (actual_time %% 1) * 100 / 60) %>%
  inner_join(tb_ref, by="trip_type") %>%
  mutate(estimated_rate_card_fare=0.75 * (base_fare + estimated_time_decimal * per_minute + estimated_distance * per_km),
         actual_rate_card_fare=0.75 * (base_fare + actual_time_decimal * per_minute + actual_distance * per_km),
         upfront_diff = upfront_offer - estimated_rate_card_fare
         ) 
# Uncomment to upload the cleaned data to the Google Sheet.
# tb %>% write_sheet(ss=ss_id, sheet="Cleaned")
nrow(tb)
```


```{r}
#| label: first-plot
#| echo: FALSE

x_max=100
y_max=x_max
distance_min = 0
distance_max = 100

tbm <- tb %>% 
  filter(trip_type=="UberX",
         date >= as.Date("2024-10-08"), 
         estimated_distance > distance_min,
         estimated_distance < distance_max,
         upfront_offer > 0,
         estimated_time_decimal > 0) 

model <- lm(upfront_diff ~ 0 + estimated_rate_card_fare, 
                      data=tbm %>% filter(estimated_rate_card_fare < 50 | upfront_diff < 0))

coeff <- model$coefficients[["estimated_rate_card_fare"]]
intercept = 0

tbp <- tbm %>%
  mutate(best_fit = estimated_rate_card_fare * coeff + intercept)

p <- ggplot(data=tbp, 
            aes(x=estimated_rate_card_fare, 
                y=upfront_diff, 
                colour="dummy"))

p + 
  geom_point() +
  # geom_smooth(method="lm", linewidth=0.5, , se=FALSE, linetype=2) +
  annotate("rect", xmin=-Inf, xmax=Inf, ymin=0, ymax=Inf, linewidth=0, fill="#1b9e77", alpha=0.15) +
  annotate("rect", xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=0, linewidth=0, fill="#d95f02", alpha=0.15) +
  annotate("label", x=0.1 * (max(tbp$estimated_rate_card_fare)) + min(tbp$estimated_rate_card_fare),
           y=0.5 * max(tbp$upfront_diff),
           label="Upfront pays more") +
  annotate("label", x=0.1 *  (max(tbp$estimated_rate_card_fare)) + min(tbp$estimated_rate_card_fare),
           y=0.7 * min(tbp$upfront_diff),
           label="Upfront pays less") +
  annotate("label", x=0.2 *  (max(tbp$estimated_rate_card_fare)) + min(tbp$estimated_rate_card_fare),
           y=0.8 * max(tbp$upfront_diff),
           label=sprintf("On average, the Upfront offer\nis only %s%% of rate card earnings*", 100 *(1 + round(coeff, 2))),
           hjust=0) +
  geom_line(aes(x=estimated_rate_card_fare, y=best_fit), colour="#7570b3") +
  labs(title=sprintf("Upfront offers and rate card estimates (%s UberX screenshots)", nrow(tbp)),
       x="Estimated rate card fare ($)",
       y="Change from rate card to Upfront ($)",
      caption="Estimated rate card fare = 0.75 * ($0.18/min + $0.81/km + $3.17).\n* Fit excludes outliers with rate card > 50 and upfront change > 0.") + 
  theme(legend.position="none",
        plot.title=element_text(hjust=0),
        plot.subtitle=element_text(hjust=0),
        plot.caption = element_text(hjust=0)) +
  scale_colour_brewer(palette = "Dark2") 
  # scale_colour_brewer(palette="BuPu") 
```

---

```{r}
#| label: model-histogram
#| echo: FALSE

x_max=100
y_max=x_max
distance_min = 0
distance_max = 100

tbm <- tb %>% 
  filter(trip_type=="UberX",
         date >= as.Date("2024-10-08"), 
         estimated_distance > distance_min,
         estimated_distance < distance_max,
         upfront_offer > 0,
         estimated_time_decimal > 0) 

model <- lm(upfront_diff ~ 0 + estimated_rate_card_fare, 
                      data=tbm %>% filter(estimated_rate_card_fare < 50 | upfront_diff < 0))

coeff <- model$coefficients[["estimated_rate_card_fare"]]
intercept = 0

tbp <- tbm %>%
  mutate("best_fit"= estimated_rate_card_fare * coeff + intercept,
         "upfront_over_estimated" = upfront_offer / estimated_rate_card_fare)

p <- ggplot(data=tbp, 
            aes(x=upfront_over_estimated),
            fill="dummy")

p + 
  # geom_density(binwidth=0.05, width=0.8, alpha=0.5) +
  annotate("rect", xmin=1, xmax=Inf, ymin=-Inf, ymax=Inf, linewidth=0, fill="#1b9e77", alpha=0.15) +
  annotate("rect", xmin=-Inf, xmax=1, ymin=-Inf, ymax=Inf, linewidth=0, fill="#d95f02", alpha=0.15) +
  annotate("label", x=0.8 * max(tbp$upfront_over_estimated),
           y=25,
           label="Upfront\npays more") +
  annotate("label", x=0.1 * min(tbp$upfront_over_estimated),
           y=25, hjust=0,
           label="Upfront\npays less") +
  geom_histogram(binwidth=0.05, width=0.04, colour="white", fill="#1b9e77") +
  geom_vline(xintercept=1.0 + coeff, colour="#7570b3") +  # colour="#7570b3") +
  annotate("label", x=1.0 + coeff,
           y=45,
           label=sprintf("On average, the Upfront offer\nis only %s%% of rate card earnings*", 100 *(1 + round(coeff, 2))),
           hjust=-0.1) +
  labs(title=sprintf("Upfront offers and rate card estimates (%s UberX screenshots)", nrow(tbp)),
       x="Upfront / rate card",
       y="Number of trips",
      caption="Estimated rate card fare = 0.75 * ($0.18/min + $0.81/km + $3.17).\n* Fit excludes outliers with rate card > 50 and upfront change > 0.",
      fill="Trip Length") +
  theme(legend.position = "inside", 
        legend.position.inside=c(0.8, 0.8),
        plot.title=element_text(hjust=0),
        plot.subtitle=element_text(hjust=0),
        plot.caption = element_text(hjust=0)) +
  #geom_vline(xintercept = 1, colour="darkgrey") +
  scale_colour_brewer(palette = "Dark2") +
  scale_fill_brewer(palette="Dark2")
```

---

```{r}
#| label: distance-and-fares
#| echo: FALSE

x_max=100
y_max=x_max

tbp <-  tb %>% 
  filter(trip_type=="UberX", 
         date >= as.Date("2024-10-08"),
         estimated_distance > 0, 
         estimated_time_decimal > 0,
         upfront_offer > 0) %>%
  filter(estimated_rate_card_fare > 0,
         estimated_rate_card_fare < x_max)

p <- ggplot(data=tbp, 
            aes(x=estimated_distance, 
                y=upfront_diff, 
                colour="dummy",
                label=comments))

p + 
  geom_point() +
  # geom_smooth(method=lm) +
  # geom_text(colour="black", size=3) +
  # geom_smooth(method="lm", linewidth=0.5, linetype=2, se=FALSE) +
  # geom_hline(yintercept=0.0, colour="darkgrey") +
  annotate("rect", xmin=-Inf, xmax=Inf, ymin=0, ymax=Inf, linewidth=0, fill="#1b9e77", alpha=0.15) +
  annotate("rect", xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=0, linewidth=0, fill="#d95f02", alpha=0.15) +
  annotate("label", x=0.1 * max(tbp$estimated_distance),
           y=0.8 * max(tbp$upfront_diff),
           label="Upfront pays more") +
  annotate("label", x=0.1 * max(tbp$estimated_distance),
           y=0.7 * min(tbp$upfront_diff),
           label="Upfront pays less") +
  annotate("label", x=0.5 * max(tbp$estimated_distance),
           y=0.9 * max(tbp$upfront_diff),
           label="Upfront offers are worse for long trips",
           hjust=0.5) +
  labs(title=sprintf("Upfront offers and trip distances (%s UberX screenshots)", nrow(tbp)),
       x="Estimated distance (km)",
       y="Change from rate card to Upfront ($)",
      caption="Estimated rate card fare = 0.75 * ($0.18/min + $0.81/km + $3.17)\nPoints below the grey line show upfront offers lower than rate card would give.") + 
  #scale_x_continuous(limits=c(0, x_max)) +
  #scale_y_continuous(limits=c(0, NA)) +
  theme(legend.position="none",
        plot.title=element_text(hjust=0),
        plot.subtitle=element_text(hjust=0),
        plot.caption = element_text(hjust=0)) +
  scale_colour_brewer(palette = "Dark2") 
  # scale_colour_brewer(palette="BuPu") 
```

---

```{r}
#| label: plot-zoom
#| echo: FALSE

x_max=20
y_max=x_max

tbp <-  tb %>% 
  filter(trip_type=="UberX", 
         date >= as.Date("2024-10-08"),
         estimated_distance > 0, 
         estimated_time_decimal > 0,
         estimated_distance < x_max,
         upfront_offer > 0) %>%
  filter(estimated_rate_card_fare < x_max)

p <- ggplot(data=tbp, 
            aes(x=estimated_distance, 
                y=upfront_diff, 
                colour="dummy"))

p + 
  geom_point() +
  #geom_smooth() +
  # geom_hline(yintercept=0.0, colour = "darkgrey") +
  annotate("rect", xmin=-Inf, xmax=Inf, ymin=0, ymax=Inf, linewidth=0, fill="#1b9e77", alpha=0.15) +
  annotate("rect", xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=0, linewidth=0, fill="#d95f02", alpha=0.15) +
  annotate("label", x=0.1 * max(tbp$estimated_distance),
           y=0.5 * max(tbp$upfront_diff),
           label="Upfront pays more") +
  annotate("label", x=0.1 * max(tbp$estimated_distance),
           y=0.5 * min(tbp$upfront_diff),
           label="Upfront pays less") +
  annotate("label", x=0.5 * max(tbp$estimated_distance),
           y=0.9 * max(tbp$upfront_diff),
           label="Upfront offers are sometimes better for short trips",
           hjust=0.5) +
  labs(title=sprintf("Upfront offers and rate card estimates for short trips (%s UberX screenshots)", nrow(tbp)),
       x="Estimated distance (km)",
       y="Change from rate card to Upfront ($)",
      caption="Estimated rate card fare = 0.75 * ($0.18/min + $0.81/km + $3.17)\nPoints below the grey line show upfront offers lower than rate card would give.") + 
  scale_x_continuous(limits=c(0, x_max)) +
  theme(legend.position="none",
        plot.title=element_text(hjust=0),
        plot.subtitle=element_text(hjust=0),
        plot.caption = element_text(hjust=0)) +
  scale_colour_brewer(palette = "Dark2") 
```

---


```{r}
#| label: model-bar-plot-short-long
#| echo: FALSE

x_max=400
y_max=x_max
distance_min = 1
distance_max = x_max
distance_cutoff = 10
tb_model <- tb %>% 
  filter(trip_type=="UberX",
         date >= as.Date("2024-10-08"), 
         estimated_distance > distance_min,
         estimated_distance < distance_max,
         upfront_offer > 0,
         estimated_time_decimal > 0) 
  
model <- lm(estimated_rate_card_fare ~ estimated_time + estimated_distance, data=tb_model)
model_0 <- lm(estimated_rate_card_fare ~ 0 + estimated_time + estimated_distance, data=tb_model)
model_upfront <- lm(upfront_offer ~ estimated_rate_card_fare, data=tb_model)
model_upfront_0 <- lm(upfront_offer ~ 0 + estimated_rate_card_fare, data=tb_model)

coeff_rate_card <- model_upfront$coefficients[["estimated_rate_card_fare"]]
intercept <- model_upfront$coefficients[["(Intercept)"]]
intercept = 0
tbp <- tb_model %>%
  mutate("best_fit"= estimated_rate_card_fare * coeff_rate_card + intercept,
         "upfront_over_estimated" = upfront_offer / estimated_rate_card_fare,
         "trip_length" = ifelse(estimated_distance < distance_cutoff, "less than 10km", "10km or more"))
  #pivot_longer(cols=c("upfront_offer", "best_fit"), names_to=c("category"))

p <- ggplot(data=tbp, 
            aes(x=upfront_over_estimated, 
                fill=trip_length))

p + 
  # geom_density(binwidth=0.05, width=0.8, alpha=0.5) +
  annotate("rect", xmin=1, xmax=Inf, ymin=-Inf, ymax=Inf, linewidth=0, fill="#1b9e77", alpha=0.15) +
  annotate("rect", xmin=-Inf, xmax=1, ymin=-Inf, ymax=Inf, linewidth=0, fill="#d95f02", alpha=0.15) +
  annotate("label", x=0.8 * max(tbp$upfront_over_estimated),
           y=25,
           label="Upfront\npays more") +
  geom_histogram(binwidth=0.05, position="dodge", width=1.0) +
  annotate("label", x=0.1 * min(tbp$upfront_over_estimated),
           y=25, hjust=0,
           label="Upfront\npays less") +
  labs(title=sprintf("Upfront offers and rate card estimates (%s UberX screenshots)", nrow(tbp)),
       subtitle=sprintf("Data set includes offers with estimated distance between %s km and %s km", distance_min, distance_max),
       x="Upfront / rate card",
       y="Number of trips",
      caption="Splitting the trips into two groups: longer and shorter than 10km shows how the groups are treated differently.\n10km may not be an exact cutoff.",
      fill="Trip Length") +
  theme(legend.position = "inside", 
        legend.position.inside=c(0.8, 0.8),
        plot.title=element_text(hjust=0),
        plot.subtitle=element_text(hjust=0),
        plot.caption = element_text(hjust=0)) +
  #geom_vline(xintercept = 1, colour="darkgrey") +
  scale_x_continuous(limits=c(0, NA)) +
  scale_colour_brewer(palette = "Dark2") +
  scale_fill_brewer(palette="Dark2")

```


---


```{r}
#| label: xl-plot
#| echo: FALSE

x_max=100
y_max=x_max

tbp <-  tb %>% 
  filter(trip_type=="UberXL", 
         date >= as.Date("2024-10-08"),
         estimated_distance > 0, 
         estimated_time_decimal > 0,
         upfront_offer > 0) %>%
  filter(estimated_rate_card_fare < x_max)

p <- ggplot(data=tbp, 
            aes(x=estimated_rate_card_fare, 
                y=upfront_diff, 
                colour="dummy"))

p + 
  geom_point() +
  #geom_smooth() +
  labs(title=sprintf("UberXL Upfront offers and rate card estimates (%s screenshots)", nrow(tbp)),
       x="Estimated rate card fare ($)",
       y="Change from rate card to Upfront ($)",
      caption="Estimated rate card fare = 0.75 * ($0.35/min + $1.55/km + $5.67).") + 
  annotate("rect", xmin=-Inf, xmax=Inf, ymin=0, ymax=Inf, linewidth=0, fill="#1b9e77", alpha=0.15) +
  annotate("rect", xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=0, linewidth=0, fill="#d95f02", alpha=0.15) +
  annotate("label", x=0.3 * max(tbp$estimated_rate_card_fare),
           y=0.7 * min(tbp$upfront_diff),
           label="Upfront pays less") +
  theme(legend.position="none",
        plot.title=element_text(hjust=0),
        plot.subtitle=element_text(hjust=0),
        plot.caption = element_text(hjust=0)) +
  scale_colour_brewer(palette = "Dark2") 
  # scale_colour_brewer(palette="BuPu") 
```

---


```{r}
#| label: trip-radar
#| echo: FALSE

x_max=100
y_max=x_max
tbp <- tb %>% 
  filter(trip_type=="UberX",
         date >= as.Date("2024-10-08"),
         estimated_time_decimal > 0,
         upfront_offer > 0,
         upfront_offer < y_max,
         estimated_time < x_max) %>%
  mutate(trip_radar = ifelse((estimated_distance == 0 | is.na(estimated_distance)), "Trip Radar", "Regular"),
         size = ifelse(trip_radar=="Trip Radar", 1, 0)) %>%
  arrange(trip_radar)

p <- ggplot(data=tbp, 
            aes(x=estimated_time, 
                y=upfront_offer,
                colour=trip_radar,
                size=as.factor(size)))

p + 
  geom_point(size=1*(1 + tbp$size), alpha=(0.5 + tbp$size)) +
  # geom_smooth(method="lm", linetype=3) +
  labs(title=sprintf("Regular and Trip Radar upfront offers (%s screenshots)", nrow(tbp)),
       caption="Trip Radar offers do not include an estimated distance.\nTrip Radar offers do not appear to be lower than regular Upfront offers.",
       x="Estimated time (mins)",
       y="Upfront fare offer ($)") +
  scale_x_continuous(limits=c(0, x_max)) +
  scale_y_continuous(limits=c(0, y_max)) +
  theme(legend.position="inside",
        legend.position.inside = c(0.2,0.8),
        legend.title=element_blank(),
        plot.title=element_text(hjust=0),
        plot.subtitle=element_text(hjust=0),
        plot.caption = element_text(hjust=0)) +
  scale_colour_brewer(palette = "Dark2")
```

